Solar wind, F10.7, and geomagnetic activity relationship to the equatorial plasma mass density at geosynchronous orbit

### 2016-10-13 paper/paper.tex
Edits based on review

\documentclass[12pt]{article}
\bibliographystyle{./style/agufull}
\usepackage[square,comma,authoryear]{natbib}
\usepackage{aas_macros}
\usepackage{epsfig}
\usepackage{amsmath}
\usepackage{style/lineno}
%\usepackage{booktabs}
\linenumbers*[1]
\usepackage{subfig}

@@

\pagestyle{fancy}
\fancyhead{}
\fancyfoot{}
\fancyhead[R]{X - \thepage}
\fancyhead[L]{VEIBELL: MASS DENSITY AT GEO}
\fancyfoot[L]{D R A F T}
\fancyfoot[C]{\today}
\fancyfoot[R]{D R A F T}
\renewcommand{\headrulewidth}{0.0pt}
\renewcommand{\footrulewidth}{0.0pt}

\oddsidemargin 0.0in
\evensidemargin 1.0in
\textwidth 6.5in
\headheight 0.1in
\topmargin -0.5in
\textheight 9.0in

@@

\begin{abstract}
We consider two types of events, identified by decreases in $D_{st}$ below a threshold value and increases in the the equatorial mass density at geosynchronous altitudes, $\rho_{eq}$, above a threshold value using the \cite{Takahashi2010} dataset.  From the $D_{st}$ events and 1-day averages, we find that there is a statistically weak and small-amplitude difference between $\rho_{eq}$ on the day of the event and the days before and after.  When hourly averages are considered, a significant peak is found to occur six hours after event onset, and the primary factor that determines the post-onset peak amplitude in $\rho_{eq}$ is elevated $F10.7$.  In addition, for hourly averages, $\rho_{eq}$ following the onset of a $D_{st}$ event depends on the north-south component of the interplanetary magnetic field, $B_z$ after the time of onset, with higher average $B_{z}$ four hours after the event onset corresponding to larger $\rho_{eq}$ value 7-11 hours after onset.  From the $\rho_{eq}$ events, we find a weak dependence on $B_{z}$ after the onset of an event, with higher average $B_{z}$ four hours after the event onset corresponding to larger $\rho_{eq}$ values 24-36 hours after onset.
\end{abstract}
%\vspace{2em}
%\hrule
%\vspace{2em}
% \end{@twocolumnfalse}
%  ]

@@

\cite{Takahashi2010} developed a mass density dataset using measurements from the Space Environment Monitor instruments on the Geostationary Operational Environmental Satellites (GOES) satellites from 1980 through 1992, with most measurements in the range of $L=6.8\pm0.2 R_E$. The mass density was estimated using the the Alfv\'en wave velocity relationship, $V_A=B/\sqrt{\mu_0\rho}$, a magnetic field model, and a numerical solution to a wave equation with an ionospheric boundary and the assumption of a zero resistance ionosphere.  The equatorial mass density, $\rho_{eq}$, was derived from the estimated mass density using a power law dependence on the geocentric distance to the field line of the observation, $R$, $\rho=\rho_{eq}(LR_E/R)^{1/2}$.  They found a high correlation ($\sim 0.94$) between 27-day averages of $F10.7$ and 27-day medians of $\rho_{eq}$.

@@

Our first analysis uses GOES-6 data only from the time interval 1989-1991, which corresponds to the interval used in the epoch analysis of \cite{Takahashi2010}. In this interval, there were 75 $D_{st}$ events where $D_{st}$ remained below the $-50$~nT threshold for at least 12 hours. Only events with an onset time between 06:00 and 12:00 MLT were considered. The first hour of a local minimum in $D_{st}$ following a threshold crossing defines the zero epoch time, and for each event, $4\cdot24$ hours were considered before and $8\cdot24$ hours after onset. For multiple detected threshold crossings within 24 hours of each other, the first crossing was selected as the event onset because visual inspection showed the subsequent crossings to typically be from brief downward deviations during the recovery period of a geomagnetic storm. To compute the epoch averages on a daily time scale, the median value of all available measurements for all events centered on a window of $\pm 12$ hours of the epoch zero hour was computed, and these averaging windows were shifted in increments of $24$~hours. Similar results are obtained if we first reduce each event time series to have a 1-day cadence by computing medians for each event in 1-day bins and then compute the medians across the averaged event time series on each epoch day.

@@

The process for creating Figure~\ref{fig:BzBinnedDstEvents} was repeated except using the sort variable $F10.7$ instead of $B_z$, and the result is shown in Figure~\ref{fig:F107BinnedDstEvents}. The difference between the epoch curves is statistically significant at all times, as expected due to the high correlation between $\rho_{eq}$ and $F10.7$, but elevated $F10.7$ is associated with a peak in the $\rho_{eq}$ curve approximately 5 hours after onset. The values contributing to the peak were compared to a baseline of all values 16-48 hours after onset via a Wilcoxon rank-sum test and were found to have significantly different medians at the 99\% confidence level.  This indicates the possibility that there must be a high pre-existing baseline level of $\rho_{eq}$ in order for an increase in $\rho_{eq}$ to be observed for a $D_{st}$ event.  This result is also consistent with the the analysis of Figure~3 when repeated using measurements from GOES satellites taken during different parts of the solar cycle; the significance of the differences between the days around onset were largest when the measurements were taken during time intervals when F10.7 was consistently elevated.

@@

Figure~\ref{fig:BzBinned} shows the the events of Figure~\ref{fig:HourlyAveragedRhoEvents} when separated by $B_z$ in the same manner as Figure~\ref{fig:BzBinnedDstEvents}.  The largest difference occurs when the separation is performed after onset, with more positive $B_z$ leading to a larger peak in $\rho_{eq}$ than more negative $B_{z}$.  The rate of change of $\rho_{eq}$ also appears to be insensitive to the time windows used to determine the separation (before or after onset).

@@

On average, around the time of large geomagnetic events, there is a statistically weak enhancement of daily-averaged $\rho_{eq}$ measurements. For hourly averages, $\rho_{eq}$ shows enhancements over a 6-hour post-onset window that contains too few events for a statistical analysis.  When the constraint of extended geomagnetic events (greater than 12 hours below a threshold value) and the MLT of the observation of $\rho_{eq}$ (06:00-12:00) is removed to allow for more events, statistically significant variations are found, with a peak occurring approximately six hours after the event onset.  For the hourly-averaged measurements, $\rho_{eq}$ averages within 7-11 hours after the onset time was lower when the average of $B_{z}$ four hours after the onset time was more positive.  This result highlights the complexity of the response of $\rho_{eq}$ to solar wind driving.  Negative values of $B_{z}$ are generally needed for a large geomagnetic event to occur and we have shown an association between large geomagnetic events and increases in $\rho_{eq}$ within six hours of the event onset, but after the onset of the event, more positive values of $B_z$ are associated with larger $\rho_{eq}$ values.

@@

Work at Dartmouth College was supported by NSF grant AGS-1105790

>>>>

\documentclass[12pt]{article}
\bibliographystyle{./style/agufull}
\usepackage[square,comma,authoryear]{natbib}
\usepackage{aas_macros}
\usepackage{epsfig}
\usepackage{amsmath}
\usepackage{style/lineno}
\usepackage[hidelinks]{hyperref}
%\usepackage{url}

%\usepackage{booktabs}
\linenumbers*[1]
\usepackage{subfig}

@@

\pagestyle{fancy}
\fancyhead{}
\fancyfoot{}
\fancyhead[R]{X - \thepage}
\fancyhead[L]{VEIBELL: MASS DENSITY AT GEO}
\fancyfoot[L]{D R A F T}
\fancyfoot[C]{\today}
\fancyfoot[R]{D R A F T}
\renewcommand{\headrulewidth}{0.0pt}
\renewcommand{\footrulewidth}{0.0pt}
\oddsidemargin 0.0in
\evensidemargin 1.0in
\textwidth 6.5in
\headheight 0.1in
\topmargin -0.5in
\textheight 9.0in

@@

\begin{abstract}
We consider two types of events, identified by decreases in $D_{st}$ below a threshold value and increases in the equatorial mass density at geosynchronous altitudes, $\rho_{eq}$, above a threshold value using the \cite{Takahashi2010} dataset.  From the $D_{st}$ events and 1-day averages, we find that there is a statistically weak and small-amplitude difference between $\rho_{eq}$ on the day of the event and the days before and after.  When hourly averages are considered, a significant peak is found to occur six hours after event onset, and the primary factor that determines the post-onset peak amplitude in $\rho_{eq}$ is elevated $F10.7$.  In addition, for hourly averages, $\rho_{eq}$ following the onset of a $D_{st}$ event depends on the north-south component of the interplanetary magnetic field, $B_z$ after the time of onset, with higher average $B_{z}$ four hours after the event onset corresponding to larger $\rho_{eq}$ value 7-11 hours after onset.  From the $\rho_{eq}$ events, we find a weak dependence on $B_{z}$ after the onset of an event, with higher average $B_{z}$ four hours after the event onset corresponding to larger $\rho_{eq}$ values 24-36 hours after onset.
\end{abstract}
%\vspace{2em}
%\hrule
%\vspace{2em}
% \end{@twocolumnfalse}
%  ]

@@

\cite{Takahashi2010} developed a mass density dataset using measurements from the Space Environment Monitor instruments on the Geostationary Operational Environmental Satellites (GOES) satellites from 1980 through 1992, with most measurements in the range of $L=6.8\pm0.2 R_E$. The mass density was estimated using the Alfv\'en wave velocity relationship, $V_A=B/\sqrt{\mu_0\rho}$, a magnetic field model, and a numerical solution to a wave equation with an ionospheric boundary and the assumption of a zero resistance ionosphere.  The equatorial mass density, $\rho_{eq}$, was derived from the estimated mass density using a power law dependence on the geocentric distance to the field line of the observation, $R$, $\rho=\rho_{eq}(LR_E/R)^{1/2}$.  They found a high correlation ($\sim 0.94$) between 27-day averages of $F10.7$ and 27-day medians of $\rho_{eq}$.

@@

Our first analysis uses GOES-6 data only from the time interval 1989-1991, which corresponds to the interval used in the epoch analysis of \cite{Takahashi2010}. In this interval, there were 75 $D_{st}$ events where $D_{st}$ remained below the $-50$~nT threshold for at least 12 hours. Only events with an onset time when the spacecraft was located between 06:00 and 12:00 MLT were considered. The first hour of a local minimum in $D_{st}$ following a threshold crossing defines the zero epoch time, and for each event, $4\cdot24$ hours were considered before and $8\cdot24$ hours after onset. For multiple detected threshold crossings within 24 hours of each other, the first crossing was selected as the event onset because visual inspection showed the subsequent crossings to typically be from brief downward deviations during the recovery period of a geomagnetic storm. To compute the epoch averages on a daily time scale, the median value of all available measurements for all events centered on a window of $\pm 12$ hours of the epoch zero hour was computed, and these averaging windows were shifted in increments of $24$~hours. Similar results are obtained if we first reduce each event time series to have a 1-day cadence by computing medians for each event in 1-day bins and then compute the medians across the averaged event time series on each epoch day.

@@

The process for creating Figure~\ref{fig:BzBinnedDstEvents} was repeated except using the sort variable $F10.7$ instead of $B_z$, and the result is shown in Figure~\ref{fig:F107BinnedDstEvents}. The difference between the epoch curves is statistically significant at all times, as expected due to the high correlation between $\rho_{eq}$ and $F10.7$, but elevated $F10.7$ is associated with a peak in the $\rho_{eq}$ curve approximately 5 hours after onset. The values contributing to the peak were compared to a baseline of all values 16-48 hours after onset via a Wilcoxon rank-sum test and were found to have significantly different medians at the 99\% confidence level.  This indicates the possibility that there must be a high pre-existing baseline level of $\rho_{eq}$ in order for an increase in $\rho_{eq}$ to be observed for a $D_{st}$ event.  This result is also consistent with the analysis of Figure~3 when repeated using measurements from GOES satellites taken during different parts of the solar cycle; the significance of the differences between the days around onset were largest when the measurements were taken during time intervals when F10.7 was consistently elevated.

@@

Figure~\ref{fig:BzBinned} shows the events of Figure~\ref{fig:HourlyAveragedRhoEvents} when separated by $B_z$ in the same manner as Figure~\ref{fig:BzBinnedDstEvents}.  The largest difference occurs when the separation is performed after onset, with more positive $B_z$ leading to a larger peak in $\rho_{eq}$ than more negative $B_{z}$.  The rate of change of $\rho_{eq}$ also appears to be insensitive to the time windows used to determine the separation (before or after onset).

@@

On average, around the time of large geomagnetic events, there is a statistically weak enhancement of daily-averaged $\rho_{eq}$ measurements. For hourly averages, $\rho_{eq}$ shows enhancements over a 6-hour post-onset window that contains too few events for a statistical analysis.  When the constraint of extended geomagnetic events (greater than 12 hours below a threshold value) and the spacecraft MLT for observation of $\rho_{eq}$ (06:00-12:00) is removed to allow for more events, statistically significant variations are found, with a peak occurring approximately six hours after the event onset.  For the hourly-averaged measurements, $\rho_{eq}$ averages within 7-11 hours after the onset time was lower when the average of $B_{z}$ four hours after the onset time was more positive.  This result highlights the complexity of the response of $\rho_{eq}$ to solar wind driving.  Negative values of $B_{z}$ are generally needed for a large geomagnetic event to occur and we have shown an association between large geomagnetic events and increases in $\rho_{eq}$ within six hours of the event onset, but after the onset of the event, more positive values of $B_z$ are associated with larger $\rho_{eq}$ values.

@@

Work at Dartmouth College was supported by NSF grant AGS-1105790.

The mass density data used for this work can be obtained from \url{http://www.dartmouth.edu/~rdenton} and the gap-filled solar wind data is available at \url{http://virbo.org/ftp/users/kondrashov/}.
